---
title: "Adding Instance Number"
author: "John Muschelli"
params:
  description: "This rundown shows how to add the instance number to a DICOM data set, using ImagePositionPatient to have the data to stack correctly."
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Description 

```{r, results = "asis", echo = FALSE}
cat(params$description)
```



# Load up Packages

```{r}
library(dcm2niir)
library(dcmtk)
library(readr)
library(dplyr)
library(tidyr)
library(neurobase)
```

# Data


The data is located at https://johnmuschelli.com/fixing_dicom_files/add_instance_number/data.zip


## Download the data

```{r}
destfile = "data.zip"
if (!file.exists(destfile)) {
  url = "https://johnmuschelli.com/fixing_dicom_files/add_instance_number/data.zip"
  destfile = tempfile(fileext = ".zip")
  download.file(url, destfile)
}
```

My version of `dcm2niix` is the most up to date GitHub built from source:

```{r}
dcm2nii_version()
```

If not, you can run `install_dcm2nii(from_source = TRUE, overwrite = TRUE)`. 

Here we are unzipping the data to a temporary file so we can sandbox our changes to the data:

```{r}
exdir = tempfile()
dir.create(exdir)
unzip(destfile, exdir = exdir)
paths = list.files(pattern = ".dcm", path = exdir, full.names = TRUE)
```

Here is the header information:

```{r}
# showing the image position patient
hdr = read_dicom_header(path = exdir, recursive = TRUE)
wide = wide_hdr(hdr)

sort(unique(hdr$name))
```

## Helper Functions

In order to parse some information on the DICOM header, we need some helper functions:

```{r}
library(readr)
parse_number_no_na = function(x) {
  x[ x %in% c("-", "-", "N/A")] = NA
  na_x = is.na(x)
  x = readr::parse_number(x, na = "")
  bad = is.na(x) & !na_x
  stopifnot(!any(bad))
  x
}

sub_bracket = function(x) {
  x = sub("^\\[", "", x)
  x = sub("\\]$", "", x)
  x = trimws(x)
}
```

## DICOM Headers are missing

```{r}
# no gantry tilt DICOM header!
any(grepl("ilt", tolower(hdr$name)))
any(grepl("instancenum", tolower(hdr$name)))

wide = wide %>% 
  mutate(ipp = sub_bracket(gsub("\\\\", ",", ImagePositionPatient))) %>% 
  tidyr::separate(ipp, into = c("x", "y", "z"), remove = FALSE, sep = ",") %>% 
  mutate_at(vars(x, y, z), parse_number_no_na)

# order by z
wide = wide %>% 
  select(file, x, y, z, ipp) %>% 
  arrange(z)
```

Showing the image position patients (ordered for instance number later)

```{r}
wide %>% 
  mutate(file = basename(file)) %>% 
  select(file, x, y, z) %>% 
  as.data.frame
```

Here is the output of `dcm2niix` on the original data

```{r out1}
d_res = dcm2nii(
  basedir = exdir,
  opts = " -v y ")
d_res
img = readnii(d_res$nii_after)
dim(img)
ortho2(window_img(img))
```


Let's look at the last slice:

```{r last_slice}
ortho2(window_img(img), xyz = c(202, 215, 1))
```

# Fixing the Instance Number

## Copy the data

Here we are copying the data to a different folder to make instance number

```{r}
tmpdir = tempfile()
dir.create(tmpdir)
file.copy(paths, tmpdir)
```

## Modifying the Instance Number

Here we are making the instance number spatial on z, modifying the instance number using the tag `(0020,0013)`: 

```{r}
tmp_paths = file.path(tmpdir, basename(wide$file))
ind = seq_along(tmp_paths)
add_instance = function(file, index) {
  dcmtk::dcmodify(
    file = file,
    frontopts = paste0('-i "(0020,0013)=', index, '"')
  )
  hdr = read_dicom_header(file)
  new_inst = as.numeric(sub_bracket(hdr$value[hdr$name == "InstanceNumber"]))
  stopifnot(new_inst == index)
  print(new_inst)
  bakfile = paste0(file, ".bak")
  if (file.exists(bakfile)) {
    file.remove(bakfile)
  }
}
res = mapply(add_instance, tmp_paths, ind)
```

Checking the instance number is correct:

```{r}
tmp_hdr = lapply(tmp_paths, read_dicom_header)
tmp_hdr = bind_rows(tmp_hdr)
# arranging because reshaping can cause reordering of files
tmp_hdr = tmp_hdr %>% 
  mutate(file = factor(tmp_hdr$file, levels = tmp_paths)) %>% 
  arrange(file)
tmp_wide = wide_hdr(tmp_hdr)
tmp_wide$InstanceNumber
```

Checking the output with filled in instance number

```{r}
inst_res = dcm2nii(
  basedir = tmpdir,
  # files = tmp_paths,
  opts = " -v y ")
inst_res
inst_img = readnii(inst_res$nii_after)
dim(inst_img)
```

## Correct Output 

Note the last slice:

```{r correct_output}
ortho2(window_img(inst_img))
```


Let's look at the last slice:

```{r, last_output}
ortho2(window_img(inst_img), xyz = c(202, 215, 1))
```


<details>

<summary>Session info</summary>

Here is my session information to ensure reproducibility for packages:

```{r}
devtools::session_info()
```

</details>
